The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-25 manuscript/rebuttal.Rmd

* Key points

### R2 Point 5

> 5. Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial. Instead of
> merely summarizing the recommendations, it would be useful to indicate the potential trajectory of
> computational research if these recommendations are adopted, as well as highlighting the potential
> risks if they aren't.

Response: We have completely rewritten the conclusion.
Instead of reiterating the five pillars, we provide the following:

> As the chemotherapeutics case study highlights, there are significant risks to poorly conducted
> bioinformatics research, and current standards are falling short.
> If adopted widely, the transparency and reproducibilty measures proposed here will reduce the chance of
> such disasters from happening again.
> Researchers, their institutions, publishers and funders each have a major role to play in fighting the
> reproducibility crisis by encouraging highly reproducible research practices.

@@

## Reviewer 3

### R3 Point 1

> The paper is mostly well written and addresses an important topic: computational reproducibility and
> tools to achieve it. The authors describe five "pillars" under which various tools and techniques can
> be categorized. The paper is pretty comprehensive in what it covers and has useful supplementary
> material. My main issue with the paper is that it reiterates many things that have been covered in
> other papers. The authors cite 11 such papers. Many of the topics covered in submitted paper have
> already been covered well in other papers. I am most familiar with reference 13, which covers many of
> the same topics, although this paper goes into more detail on many issues. The authors could do more
> to differentiate their paper from previous ones and perhaps remove some topics that are already
> covered well elsewhere.

RESPONSE: We thank reviewer 3 for their careful assessment of our article and appreciate the suggestions
which have improved the manuscript.
Indeed, this article does reiterate a great deal of what has already been said, it is a review after
all.
We note that each of those recommendation articles mentioned have different emphases.
For example, some of them don't speak to the benefits of literate programming,
while others do not mention the importance of documentation.
Some are focused on specific concepts like infrastructure, containerisation or
continuous analysis.
Some are proof of principle workflows, which present a rigid combination of tools.
Here, we describe the ways computational researchers can make their own work more reproducible.
We are assuming that the readers are not advanced analysts, so some topics like pipelining systems
are covered in less detail and we refer the reader to those relevant reviews.
What does make this article unique is the Five Pillars infographic, and specific combination of
recommendations.
Our article also provides a handy list of recommended tutorials as a supplementary resource.

@@

RESPONSE: Citations have been added for Podman and Apptainer/Singularity.
Rkt was removed, as it appears that development of the project has ceased.

### R3 Point 16

> 16. Page 9, lines 7-8: I believe you, but I am not aware of evidence that supports this claim.

RESPONSE: The claim made in the manuscript is that "Docker remains the most widely used
containerisation system for web development and research".
We do not have exact usage data on this, however we can look at other metrics like GitHub stars.

| Container System | GitHub stars* |
| --- | --- |
| Docker | 72.2k |
| Podman | 18.9k |
| Singularity | 3.1k |

* 25th August.

Based on this we have altered the text:

> While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
> [@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
> stars).

@@

> 5. Use file formats that are machine-readable and compatible with many different types of software.
> For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
> specifications that are suitable for very large data.

### R3 Point 25

> 25. One thing that could be added is something about Common Workflow Language.
It's a community-supported specification for accomplishing many of the objectives described here.
There are some recent papers about this.

RESPONSE: Thanks for raising this. We have added CWL to the section mentioning pipeline solutions, and
refered readers to dedicated reviews on the topic.

> Projects involving computationally intensive tasks would benefit from a build/workflow automation
> solution.
> Guides for selecting and using such systems have been already reviewed by others [@Leipzig2016-bf;
> @Reiter2021-lh].
> But it is worth mentioning some of the most used solutions in bioinformatics, which include
> `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021], `CWL` [@Crusoe2022-cb],
> `WDL` [@Voss2017-vd] and `nextflow` [@Di_Tommaso2017-yo].
> The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
> for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
> need to be repeated.
> After fixing the issue, re-running the analysis would pick up at step 8 again,
> which saves labor and compute time.

@@

### R3 Point 26

> 26. Page 14, lines 9-10: The big question is how to make more progress.
> We have the tools to achieve reproducibility, but why are we rarely achieving it?
> The paper mentions incentives and lack of training, which are true.
> You might consider elaborating a bit.
> My cynical view is that writing more papers and tutorials will do little without strong incentives
> and more automation.

RESPONSE: Thanks for raising this, and I note that similar sentiments were shared by reviewer 2 about the
challenges that are holding back change.
Please refer to our response to Reviewer 2's Point #3.
The Challenges section now includes brief mentions of incentives and automation.

### 2023-08-25 manuscript/5pillars.Rmd
rev3 done

library("RColorBrewer")
cols <- brewer.pal(5, "Set2")
cols2 <- adjustcolor( cols, alpha.f = 0.7)
plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 9))
polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
polygon(c(2,3.8,3.8,2),c(1,1,6,6), col=cols2[2], border = NA)
polygon(c(4,5.8,5.8,4),c(1,1,6,6), col=cols2[3], border = NA)
polygon(c(6,7.8,7.8,6),c(1,1,6,6), col=cols2[4], border = NA)
polygon(c(8,9.8,9.8,8),c(1,1,6,6), col=cols2[5], border = NA)
polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
polygon(c(-0.2,5,10),c(6.3,8.5,6.3),col="lightgray", border = NA)
points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
text(5,9,lab="Five pillars of reproducible computational research", cex=1.5)

@@

A guiding principle of the five pillars approach is that the publishing researchers should strive
to make the reproduction process easier and faster for those who wish to.
A "master script" that coordinates the execution of individual parts of an analysis is an excellent
way to reduce the number of commands required for reproduction.
This can be coded in standard open source languages like R/R Markdown, Python/Jupyter notebook or the
Unix Shell.
These approaches work equally well on personal computers, cloud servers and high performance clusters.
Using free and open source software ensures that the research is accessible to the greatest audience
@Garijo2013-zl, as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
the scientific community.
Projects involving computationally intensive tasks might benefit from a build/workflow automation
solution like `make`, `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
`nextflow` [@Di_Tommaso2017-yo].
The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
need to be repeated.
After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
which saves labor and compute time.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
[@Kurtzer2017-zq], Docker remains the most widely used containerisation system for web development and research.
Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the codebase and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.

@@

@@

Due to the relative lack of systematic reproducibility/replication studies in bioinformatics,
the severity of the reproducibility problem isn't well appreciated.
The studies that have been conducted point to reproducibility rates of ~10% in computational
life sciences [@Ioannidis2009-hv;@Zaringhalam2020-ms; @Samuel2022-ha], but further systematic studies
are required to bring attention to the issue.

Journals are partly responsible, as their editorial policies influence author behaviours.
Except for a few outliers, verification of computational reproducibility is not in the scope of peer
review, but this should be considered.
As others have noted, it is in the interests of funders to provide the infrastructure to facilitate this
[@Peng2011-ia].

Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].
This is a direct result of research assessment policies that incentivize journal-level prestige
metrics and number of publications rather than the rigour [@Munafo2017-gf].
A change in research assessment ("eg: track-record") that recognises and credits contributions toward
reproducibility and rigor would help decrease funds wasted on sloppy research.

@@

## Future directions

@@

Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
significant expertise to analyze and distill key observations.
The measures proposed here are designed to enhance transparency and foster the reuse of data and code;
reducing research waste.
Transparency makes it easier for problems to be identified, so that erroneous findings are less likely to
pollute the scientific literature.
Researchers, their institutions, publishers and funders each have a major role to play in fighting the
reproducibility crisis by encouraging highly reproducible research practices.

>>>>

library("RColorBrewer")
cols <- brewer.pal(5, "Set2")
cols2 <- adjustcolor( cols, alpha.f = 0.7)
plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 9.7), ylim=c(0, 9))
polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
polygon(c(2,3.8,3.8,2),c(1,1,6,6), col=cols2[2], border = NA)
polygon(c(4,5.8,5.8,4),c(1,1,6,6), col=cols2[3], border = NA)
polygon(c(6,7.8,7.8,6),c(1,1,6,6), col=cols2[4], border = NA)
polygon(c(8,9.8,9.8,8),c(1,1,6,6), col=cols2[5], border = NA)
polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
polygon(c(-0.2,4.9,10),c(6.3,8.5,6.3),col="lightgray", border = NA)
points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=16,col=cols )
text(4.8,9,lab="Five pillars of reproducible computational research", cex=1.5)

@@

A guiding principle of the five pillars approach is that the publishing researchers should strive
to make the reproduction process easier and faster for those who wish to.
A "master script" that coordinates the execution of individual parts of an analysis is an excellent
way to reduce the number of commands required for reproduction.
This can be coded in standard open source languages like R/R Markdown, Python/Jupyter notebook or the
Unix Shell.
These approaches work equally well on personal computers, cloud servers and high performance clusters.
Using free and open source software ensures that the research is accessible to the greatest audience
@Garijo2013-zl, as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
the scientific community.
Projects involving computationally intensive tasks would benefit from a build/workflow automation
solution.
Guides for selecting and using such systems have been already reviewed by others [@Leipzig2016-bf;
@Reiter2021-lh].
But it is worth mentioning some of the most used solutions in bioinformatics, which include
`snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021], `CWL` [@Crusoe2022-cb],
`WDL` [@Voss2017-vd] and `nextflow` [@Di_Tommaso2017-yo].
The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
need to be repeated.
After fixing the issue, re-running the analysis would pick up at step 8 again,
which saves labor and compute time.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
[@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
stars as of August 2023).
Docker has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the codebase and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.
A recently described set of recommendations for making software FAIR describes further measures
that data analysts could use to enhance their code documentation.

@@

### Process validation and testing

@@

Due to the relative lack of systematic reproducibility/replication studies in bioinformatics,
the severity of the reproducibility problem isn't well appreciated.
The studies that have been conducted point to reproducibility rates of ~10% in computational
life sciences [@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], but further systematic studies
are required to bring attention to the issue.

Journals are partly responsible, as their editorial policies influence author behaviours [@Han2017-pz].
Except for a few outliers, verification of computational reproducibility is not in the scope of peer
review, but automating this in the publication process should be considered.
As others have noted, it is in the interests of funders to provide the infrastructure to facilitate this
[@Peng2011-ia].

Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].
This is a direct result of research assessment policies that incentivize journal-level prestige
metrics and number of publications rather than the rigor [@Munafo2017-gf; @Oransky2023-zb].
A change in research assessment ("eg: track-record") that recognises and credits contributions toward
reproducibility and rigor would help decrease funds wasted on sloppy research.

@@

@@

As the chemotherapeutics case study highlights, there are significant risks to poorly conducted
bioinformatics research; and current standards are falling short.
If adopted widely, the transparency and reproducibilty measures proposed here will reduce the chance of
such disasters from happening again.
Researchers, their institutions, publishers and funders each have a major role to play in fighting the
reproducibility crisis by encouraging highly reproducible research practices.
